---
title: "UCLA - Regression Pt.1"
author: "lars pijnappel"
date: "04 july 2015"
output: html_document
source: http://scc.stat.ucla.edu/
---

#Regression in R - Part 1: Simple Linear Regression

>**Credits**
Following is largely taken from the 'Regression in R' training material by Denise Ferrari & Tiffany Head (UCLA Department of Statistics - [Statistical Consulting Center][scc]).

#2. Simple Linear Regression

Linear regression with a single predictor. The objective is to describe the relationship between *two variables*, say X and Y as a straight line, that is, Y is modeled as a *linear* function of X. 

X is an *explanatory* variable on the horizontal axis, whereas Y is a *response* variable on the vertical axis.

###Loading the data..
.. and take a first look at the data. 

```{r}
production <- read.table ("http://www.stat.tamu.edu/~sheather/book/docs/datasets/production.txt", header = T, sep = "")
summary(production)
plot(production)
```

The scatter plot allows one to check if the linear relationship is supported by the data.

```{r}
attach(production)
plot(RunTime~RunSize)
```

    A brief explanation on regression: see the reg_1.pdf (sheet 24-26).

###WIP ...



[scc]: http://scc.stat.ucla.edu/ 'UCLA Statistical Consulting Center'
