---
title: "UCLA - Regression Pt.1"
author: "lars pijnappel"
date: "04 july 2015"
output: html_document
source: http://scc.stat.ucla.edu/
---

#Regression in R - Part 1: Simple Linear Regression

>**Credits**
Following is largely taken from the 'Regression in R' training material by Denise Ferrari & Tiffany Head (UCLA Department of Statistics - [Statistical Consulting Center][scc]).

#1. Introduction

###Loading the Data

First retrieve the dataset 'Diabetes in Pima Indian Women' from the UCI Machine Learning Repository and assign it to the `pima` data frame. As this dataset contains non-descriptive columnnames, the header is set to false to discard them. Then, by using the function `colnames()`, more meaningful labels are added to this dataset.

```{r}
pima <- read.table("http://archive.ics.uci.edu/ml/machine-learning-databases/pima-indians-diabetes/pima-indians-diabetes.data", header=F, sep=",")

# bp=blood pressure / bmi=body mass index
colnames(pima) <- c("npreg", "glucose", "bp", "triceps", "insulin", "bmi", "diabetes", "age", "class")
```

###Having a peek at the data

Let's get some first impressions of the `pima` data frame. By using the structure function `str()` one sees directly how many rows (in this case 768 observations) and columns (i.e. 9 variables) this dataset holds.

```{r}
str(pima)
```

TIP: Instead of using the function `str()` for a specific object, you can use `ls.str()` to see the structural details of all objects.

```{r}
# returning the first and last parts of a vector, matrix, table, data frame or function.
head(pima)
tail(pima)
```

###Numerical Summaries

Univariate summary information: Look for unusual features in the data (data-entry errors,
outliers): check, for example, min, max of each variable.

```{r}
summary(pima)

# later in this Introduction section, the (before and after preprocessing) insulin observations are compared side-by-side.
i_before <- pima$insulin  # insulin before preprocessing.
```

###Coding Missing Data

* Variable "npreg" has maximum value equal to 17
    * unusually large but not impossible
* Variables "glucose", "bp", "triceps", "insulin" and "bmi" have minimum value equal to zero
    * in this case, it seems that zero was used to code missing data

Zero should ***not*** be used to represent missing data
* it's a valid value for some of the variables
* can yield misleading results

Set the missing values coded as zero to NA:

```{r}
pima$glucose[pima$glucose == 0] <- NA
pima$bp[pima$bp == 0] <- NA
pima$triceps[pima$triceps == 0] <- NA
pima$insulin[pima$insulin == 0] <- NA
pima$bmi[pima$bmi == 0] <- NA
```

Let's verify that the minimum value for the modified variables are not zero anymore:

```{r}
summary(pima)
```

###Coding Categorical Variables
Variable "class" is categorical, not quantitative

**R code for categorical variables**

Categorical should ***not*** be coded as numerical data (it's a problem of "average zip code"). Set categorical variables coded as numerical to factor:

```{r}
pima$class <- factor(pima$class)
# view the 'Class' variable BEFORE assigning labels to the levels
levels(pima$class)
summary(pima$class)
```

```{r}
levels(pima$class) <- c("neg", "pos")
# and again, now AFTER assigning labels to the levels
levels(pima$class)
summary(pima$class)
```

###Final Coding
```{r}
summary(pima)
```

***
####Extra: comparing the insulin observations before & after the preprocessing
Remember that for several variables, such as `insulin`, the value '0' for an observation has been replaced with 'NA'. Now look at the differences:
```{r}
i_after <- pima$insulin                 # insulin after preprocessing
i_both <- data.frame(i_before, i_after) # insulin compared
summary(i_both)
```
***

###Graphical Summaries (Univariate)
Before starting off, let's first get a 'condensed' view of the blood pressure data by using the Stem-and-Leaf plot `stem()`.
```{r}
stem(pima$bp)
```


```{r}
# simple data plot
plot(sort(pima$bp))
# histogram
hist(pima$bp)
# density plot
plot(density(pima$bp, na.rm = TRUE))
```



###Graphical Summaries (Bivariate)

```{r}
# scatterplot
plot(triceps~bmi, pima)
# boxplot
boxplot(diabetes~class, pima)
```



[scc]: http://scc.stat.ucla.edu/ 'UCLA Statistical Consulting Center'
