---
title: "DSS - Tutorial PA2"
author: "lars pijnappel"
date: "27 juni 2015"
output: html_document
source: http://datasciencespecialization.github.io/rprog/
---

#Programming Assignment 2

***
***Prerequisite before starting this tutorial***

*If you want to follow this tutorial step-by-step in your own Rmd file, then make sure that you include the next two functions as well. These functions are taken from [rdpeng / ProgrammingAssignment2][rdpeng] (which is part of the R Programming of the [Coursera Data Science Specialization][cdss]).*

####Function makeVector()
```{r}
makeVector <- function(x = numeric()) {
    m <- NULL
    set <- function(y) {
            x <<- y
            m <<- NULL
    }
    get <- function() x
    setmean <- function(mean) m <<- mean
    getmean <- function() m
    list(set = set, get = get, setmean = setmean, getmean = getmean)
}
```

####Function cachemean()
```{r}
cachemean <- function(x, ...) {
        m <- x$getmean()
        if(!is.null(m)) {
                message("getting cached data")
                return(m)
        }
        data <- x$get()
        m <- mean(data, ...)
        x$setmean(m)
        m
}
```

```{r}
## Do a simple test to verify if both functions work as expected.
z <- makeVector()
z$set(1:4)
cachemean(z) 
```

*That's it!*

***

##Step by step to understand the instructions
By [DanieleP][PA2]

Hi everyone, Assignment 2 can be very challenging to be understood. The solution is easier than what it looks, and it's strongly connected with the example given in the instructions. However, I think it's worth spending some time to understand more about the logic behind the functions given in the instruction.

##Introduction
makeVector is a function that stores a list of functions. I had no idea it was possible. Here is an easier example of a function that stores functions:

```{r}
## function plusFunctions stores two functions:
## plustwo() : sums 2 to the given value
## plusthree() : sums 3 to the given value
plusFunctions <- function (){
    plustwo <- function(y) {
        x <- y + 2
        return(x)
    }
    plusthree <- function(y) {
        x <- y + 3
        return(x)
    }
    # the following line stores the two functions:
    list(plustwo = plustwo, plusthree = plusthree) 
}
```

To use the functions stored in the main function, you need to subset the main function. To do this, you need the name of the **main function** + **"$"** + **the name of the second function** + **(arguments)**

```{r}
## Examples from the function above:
a <- plusFunctions()
a$plustwo(5)
a$plusthree(5)
```

##makeVector()
makeVector contains 4 functions: `set`, `get`, `setmean`, `getmean`. 

###Function get()
Let's start from `get`, because it's the easiest.

```{r}
get <- function() x
```

`get` is a function that returns the vector x stored in the main function. Doesn't require any input.

```{r}
## Example
a <- makeVector(c(5,1,3))
a$get()
```

###Function set()
`set` is a function that changes the vector stored in the main function.

```{r}
set <- function(y) {
            x <<- y
            m <<- NULL
    }
```

We don't need to use this function unless we want to change the vector. `x <<- y` substitutes the vector x with y (the input) in the main function (makeVector). If it was `x <- y` it would have substitute the vector x with y only in the set function. `m <<- NULL` restores to null the value of the mean m, because the old mean of the old vector is not needed anymore. The new mean needs to be recalculated through the function cachemean.

```{r}
## Example
a <- makeVector(c(5,1,3))
a$get()
## Example
a$set(c(7,4,1,2))
a$get()
```

###Functions setmean() and getmean()
`setmean` and `getmean` are functions very similar to `set` and `get`. They don't calculate the mean, they simply store the value of the input in a variable `m` into the main function `makeVector`  `(setmean)` and return it `(getmean)`.

```{r}
setmean <- function(mean) m <<- mean
getmean <- function() m
```

This value "mean", input of `setmean`, is supposed to be the mean of the vector x. However it simply stores a value, like in the following example:

```{r}
## Example
a<- makeVector(c(1,2,3,4))
a$setmean(10)
a$getmean()
```

As you see 10 is not the mean of vector (1,2,3,4), but because we stored it with `setmean`, we got it back with `getmean`.

###Function list()
To store the 4 functions in the function makeVector, we need the function `list()`, so that when we assign `makeVector` to an object, the object has all the 4 functions.

```{r}
list(set = set, get = get, setmean = setmean, getmean = getmean)
```

##Function cachemean()
Now that makeVector is a bit more clear, let's observe the second function, `cachemean`. Input of `cachemean` is the object where `makeVector` is stored.

```{r}
## Example
a <- makeVector(c(1,2,3,4))
cachemean (a)
```

The first thing `cachemean` does is to verify the value `m`, stored previously with `getmean`, exists and is not NULL. If it exists in memory, it simply returns a `message` and the value `m`, that is supposed to be the mean, but not necessarily.

```{}
## 1st snippet of the cachemean() function
m <- x$getmean()    
if(!is.null(m)) {
    message("getting cached data")
    return(m)
data <- x$get()
m <- mean(data, ...)
x$setmean(m)
m
  }
```

If it was the case, `return(m)` would have ended the function. So everything that follows this `if()` is a sort of else {}. `data` gets the vector stored with `makeVector`, `m` calculates the mean of the vector and `x$setmean(m)` stores it in the object generated assigned with `makeVector`.

```{}
## 2nd snippet of the cachemean() function
data <- x$get()
m <- mean(data, ...)
x$setmean(m)
m
```

Hope this helps!

[rdpeng]: https://github.com/rdpeng/ProgrammingAssignment2
[cdss]: https://www.coursera.org/specialization/jhudatascience/1 'Data Science course by Johns Hopkins University'
[PA2]: https://github.com/DanieleP/PA2-clarifying_instructions 'PA2-clarifying_instructions'
