---
title: "Lectures notes"
author: "lars pijnappel"
date: "10 september 2015"
output: html_document
---

***
***Orginal stuff is coming from the GitHub repository [DataScienceSpecialization](https://github.com/DataScienceSpecialization) > Courses > [03_GettingData](https://github.com/DataScienceSpecialization/courses/tree/master/03_GettingData).***

***

# Table of Contents

* [CSV and Excel files](#CSV-and-Excel-files)
* [XML files](#xml-files)
    * [XPath](#xpath)
* [JSON](#json)
* [data.table](#data.table)
* [Cleanup file system](#cleanup-file-system)

_Note to myself: To get a sublist, press 4x space (or a multiple of it for a sub-sublist)_

***

# CSV and Excel files 
[ToC](#table-of-contents)

> 01_04_downLoadingFiles.pdf

```{r}
getwd()

if ( !file.exists( "data" )) {
        dir.create( "data" )
}
dir( )
# file.remove( "data" )
# dir()
```

### download csv
```{r}
## folder references './data' and 'data' both works..

fileURL <- "https://data.baltimorecity.gov/api/views/dz54-2aru/rows.csv?accessType=DOWNLOAD"
## w/o method argument, function results in an error when Knitting it (CTRL-SHIFT-K)
download.file( fileURL, destfile = "./data/cameras.csv", method="wininet")
list.files( "data" )
```

> 01_05_readingLocalFiles.pdf

```{r}
# read.table() expects by default tab separated data and no header included.
cameraData_csv <- read.table( "./data/cameras.csv", sep=",", header = TRUE )
head( cameraData_csv )
```

> 01_06_readingExcelFiles.pdf

### download xlsx
```{r}
if ( !file.exists("data") ) dir.create("data")

fileURL <- "https://data.baltimorecity.gov/api/views/dz54-2aru/rows.xlsx?accessType=DOWNLOAD"
## w/o mode argument, the donwloaded xlsx gets corrupted (can't be opened in Excel)
download.file( fileURL, destfile = "./data/cameras.xlsx", method="wininet", mode="wb")
list.files( "data" )


dateDownloaded <- date()
dateDownloaded
```

### Different ways reading the xlsx file
```{r}
## this library requires java (64-bit > http://www.java.com/en/download/manual.jsp )..
library( xlsx, quietly = TRUE )
cameraData_xlsx <- read.xlsx( "./data/cameras.xlsx", sheetIndex = 1, header = TRUE)

## following library doesn't require java..
library( readxl )
cameraData_rxl <- read_excel( "./data/cameras.xlsx" )

## XLConnect package has more options for writing/manipulating Excel files.
## https://cran.r-project.org/web/packages/XLConnect/vignettes/XLConnect.pdf
library( XLConnect, quietly = TRUE )
cameraData_xlc <- loadWorkbook( "./data/cameras.xlsx" )

## apparently, the rxl-object differs with both the csv- and xlsx-object
identical( cameraData_csv, cameraData_xlsx )
identical( cameraData_csv, cameraData_rxl )
identical( cameraData_xlsx, cameraData_rxl )
## same applies for the xlc-object
identical( cameraData_xlc, cameraData_csv )
identical( cameraData_xlc, cameraData_xlsx )
identical( cameraData_xlc, cameraData_rxl )


i_col <- 2:3
i_row <- 1:4
cameraData_xlsx_Subset <- read.xlsx( "./data/cameras.xlsx", sheetIndex = 1, colIndex = i_col, rowIndex = i_row )
cameraData_xlsx_Subset
```


# XML files 
[ToC](#table-of-contents)

> 01_07_readingXML.pdf

Following examples are taken from [w3schools](http://www.w3schools.com/xml/simple.xml)

### Read file into R
```{r}
library(XML)

fileURL <- "http://www.w3schools.com/xml/simple.xml"
doc <- xmlTreeParse( fileURL, useInternal=TRUE )
rootNode <- xmlRoot( doc )
xmlName( rootNode )
names( rootNode )
```

### Directly accessing parts of the XML
```{r}
rootNode[[1]]
rootNode[[1]][[1]]
```

### Programatically accessing parts of the XML
```{r}
xmlSApply( rootNode, xmlValue )
```

## XPath
[ToC](#table-of-contents)

    * /node Top level node
    * //node Node at any level
    * node[@attr-name] Node with an attribute name
    * node[@attr-name='bob'] Node with attribute name attr-name='bob'

Information from: http://www.stat.berkeley.edu/~statcur/Workshop2/Presentations/XML.pdf
```{r}
xpathSApply(rootNode,"//name",xmlValue)
xpathSApply(rootNode,"//price",xmlValue)
```

### XPath - another example (html)
http://espn.go.com/nfl/team/_/name/bal/baltimore-ravens

With htmlTreeParse you basically read the source of the webpage.
```{r}
fileUrl <- "http://espn.go.com/nfl/team/_/name/bal/baltimore-ravens"
doc <- htmlTreeParse( fileUrl, useInternal = TRUE )
scores <- xpathSApply( doc, "//div[@class='score']", xmlValue )         ## <div class="score">
teams <- xpathSApply( doc, "//li[@class='team-name']", xmlValue )       ## <li class="team-name">
scores
teams


## xmlValue: raw contents of leaf nodes (which have no children nodes)
gamemeta <- xpathSApply( doc, "//div[@class='game-meta']", xmlValue)  ## <div class="game-meta">
gamemeta
## xmlChildren: returns a list of child XMLNode objects within an XMLNode object
gamemeta <- xpathSApply( doc, "//div[@class='game-meta']", xmlChildren) 
head(gamemeta, 2)
## xmlName: element or tag name introduced in the <name ...> entry in an XML document
gamemeta <- xpathSApply( doc, "//div[@class='game-meta']", xmlName)
gamemeta
## xmlAttrs: name-value pairs of attributes
gamemeta <- xpathSApply( doc, "//div[@class='game-meta']", xmlAttrs)
gamemeta

```

### *Notes and further reading*

* Official XML tutorials [short](http://www.omegahat.org/RSXML/shortIntro.pdf), [long](http://www.omegahat.org/RSXML/Tour.pdf)
* An [outstanding guide to the XML package](http://www.stat.berkeley.edu/%7Estatcur/Workshop2/Presentations/XML.pdf)


# JSON
[ToC](#table-of-contents)

> 01_08_readingJSON.pdf

The jsonlite package requires package curl, for which [a first intro on curl](http://www.r-bloggers.com/the-curl-package-a-modern-r-interface-to-libcurl/) can be found at R-bloggers. 
```{r}
library( jsonlite, quietly = TRUE)

jsonData <- fromJSON("https://api.github.com/users/jtleek/repos")
names( jsonData )
jsonData$name
```

### Nested objects in JSON
```{r}
names(jsonData$owner)
jsonData$owner$login
```

### Writing data frames to JSON
```{r}
iris3 <- head(iris, 3)
iris3
myjson <- toJSON( iris3, pretty=TRUE)
cat(myjson)
```


### Convert back to JSON
```{r}
irisfromJSON <- fromJSON(myjson)
head( irisfromJSON )
```

### *Notes and further reading*

* [www.json.org](http://www.json.org/)
* R-bloggers on [json encoderdecoder](http://www.r-bloggers.com/new-package-jsonlite-a-smarter-json-encoderdecoder/)
* [jsonlite vignette](http://cran.r-project.org/web/packages/jsonlite/vignettes/json-mapping.pdf)


# data.table
[ToC](#table-of-contents)

> 01_09_dataTable.pdf

### Create data tables just like data frames
```{r}
library(data.table, quietly = TRUE)
# set.seed(1)
DF = data.frame(x=rnorm(9),y=rep(c("a","b","c"),each=3),z=rnorm(9))
head(DF,3)
# set.seed(1)
DT = data.table(x=rnorm(9),y=rep(c("a","b","c"),each=3),z=rnorm(9))
head(DT,3)
```

### See all the data tables in memory
```{r}
tables()
```

### Subsetting rows
```{r}
DT[2,]
DT[DT$y=="b",]
DT[c(2,3)]
```

### Column subsetting in data.table
```{r}
DT[,c(2,3)]
```

* The subsetting function is modified for data.table
* The argument you pass after the comma is called an "expression"
* In R an expression is a collection of statements enclosed in curley brackets

```{r}
{ x <- 1
  y <- 2
}
k <- { print(10); 5 }
print( k )
```

### Calculating values for variables with expressions
```{r}
DT[ , list( mean(x), sum(z) )]
DT[ , table(y) ]
```

### Adding new columns
> Be carefull, as a copy of the object will also be modified when changing values..

```{r}
DT[ , w := z^2 ]
DT2 <- DT
DT[ , y := 2 ]

head(DT,n=3)
head(DT2,n=3)
```

### Multiple operations
```{r}
DT[ , m := { tmp <- (x+z); log2(tmp+5) } ]
```

### plyr like operations
```{r}
DT[ , a := x>0 ]
DT[ , b := mean(x+w), by = a ]
```

### Special variables
`.N` An integer, length 1, containing the number of elements of a factor level 
```{r}
set.seed(123)
DT <- data.table( x = sample( letters[1:3], 1E5, TRUE ))
DT
DT[ , .N, by = x ]
```

### Keys
```{r}
DT <- data.table( x = rep( c( "a", "b", "c" ), each=100 ), y = rnorm(300) )
setkey( DT, x )
DT['a']
```

### Joins
```{r}
DT1 <- data.table( x = c( 'a', 'a', 'b', 'dt1' ), y = 1:4 )
DT1
DT2 <- data.table( x = c( 'a', 'b', 'dt2'), z = 5:7 )
DT2
setkey( DT1, x ); setkey( DT2, x )
merge( DT1, DT2 )
```

### Fast reading
```{r}
big_df <- data.frame( x = rnorm(1E6), y = rnorm(1E6) )
file <- tempfile()
write.table( big_df, file = file, row.names = FALSE, col.names = TRUE, sep = "\t", quote = FALSE )
system.time( fread( file ))
system.time( read.table( file, header = TRUE, sep = "\t" ))
```

### *Notes and further reading*
* The latest development version contains new functions like melt and dcast for data.tables
    * https://r-forge.r-project.org/scm/viewvc.php/pkg/NEWS?view=markup&root=datatable
* Here is a __list of differences between data.table and data.frame__
    * http://stackoverflow.com/questions/13618488/what-you-can-do-with-data-frame-that-you-cant-in-data-table
* Notes based on [Raphael Gottardo's notes](https://github.com/earosenthal/biost578A/blob/master/Advanced_data_manipulation-EAR.Rpres), who got them from Kevin Ushey.



***
# Cleanup file system
[ToC](#table-of-contents)

Remove the created data folder (incl. its downloaded files).

```{r}
unlink( "data" , recursive = TRUE)
dir()
```

